Molecular transport in human pial perivasculature

### 2024-11-13 paper/supplementary_main.tex
More work on supplementary description of 3D-1D
Former-commit-id: 25f33f36b50624376d77049c2fed17f7f41f272d [formerly 25f33f36b50624376d77049c2fed17f7f41f272d [formerly 581f10dae6a3590eed2499a0e0d71e5f004fbea8]]
Former-commit-id: b831c57a115c61eadc6c5a60de64c7351765073b
Former-commit-id: f16d224d3238198aa07b629ebedd87897205a6ab

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(see also \Cref{fig:intracranial_domains}). The coordinates in these
3D domains is given by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
of the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

In addition, we consider two sets of perivascular networks represented
as graphs: a periarterial network $\Lambda_a$ represented by the
(connected) centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a
perivenous network $\Lambda_v$ corresponding to the centerlines
$\Lambda^i_{(v)}$ of the veins. Recall that we represent the vascular
domains as the union of cylindrical vessels of radius $R_1^i$
surrounding the centerlines $\Lambda^i$. Moreover, we consider the
surrounding perivascular spaces as the union of annular cylinders
$\Omega^i$ of inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$
and thus of width $R_2^i - R_1^i$. We denote the outer lateral surface
of the periarterial and perivenous spaces by $\Gamma_a$ and $\Gamma_v$, respectively. We
omit the subscript $a$ or $v$ when referring to any such network,
vessel segment or perivascular outer surface. We assume that $\Gamma$
is parametrized by the coordinate $s$, and with a minor abuse of
notation, simply write $s \in \Lambda$ to represent the point ${\bm
  \lambda}(s)$ on $\Lambda$ corresponding to $s$.

@@

In this section, we describe the mathematical
model~\eqref{eq:multi_transport}, the associated definitions,
interface and boundary conditions in further detail. We refer to the
main text for explicit parameter values and give the general, abstract
form here. Recall that we model a concentration field $c = c(x, t)$
for $x \in \Omega$, $t > 0$ defined in $\Omega_{\rm PAR}$ and
$\Omega_{\rm CSF}$ separately. In each of these domains, $c$ satisfies
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma = f \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivacular networks $\Lambda_a, \Lambda_v$. To summarize,
\[
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\]
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$  if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$ in which case the interaction with $\Omega_{\rm{CSF}}$ (resp. $\Omega_{\rm{PAR}}$) is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

\fixme{Also in~\eqref{eq:3d_pde}, the notation $\overline{c}$ denotes a lateral average of the concentration over the outer perivascular surfaces, defined for each centerline $\Lambda^i$ and each point $s \in \Lambda^i$ by}
\begin{equation}
  \overline{c}(s) = \frac{1}{A(s)} \int_{\Theta(s)} c 
\end{equation}
\fixme{where $\Theta(s)$ is a cross-section of the PVS $\Omega^i$ at $s$ and $A(s)$ is the area of $\Theta(s)$.} For both the periarterial and perivenous networks ($\Lambda_a, \Lambda_v$ with outer PVS boundaries $\Gamma_a, \Gamma_v$), the (Dirac delta) term $\delta_\Gamma$ is a distribution concentrated on the outer lateral surfaces of the PVSs. \discuss{More precisely, it is defined in terms of its action on any sufficiently smooth function $v : \Omega \rightarrow \R$ and for each $\Lambda^i \subseteq \Lambda$ with PVS $\Omega^i$ and outer lateral surface $\Gamma^i = \partial \Omega^i_2$ by}
\begin{equation}
  %\xi (\overline{c} - \hat c ) \delta_\Gamma (v) = \int_{\Lambda} P \xi ( \overline{c} - \hat c ) \overline{v} .
  \delta_\Gamma (v) = \int_{\Gamma} v = \sum_{i} \int_{\Gamma^i} v = \sum_i \int_{\Lambda^i} P \, \overline{v} = \int_{\Lambda} P  \, \overline{v} .
\end{equation}
where $P = P(s)$ is defined as the perimeter of each cross-section $\Theta(s)$ for $s \in \Lambda^i$.

\mer{MER: Stopped here. Feel free to check the above. I thought the overline was defined as the cross-section perimeter average and not the cross-section average. This is more compatible with the definition of $\delta_{\Gamma}$ as well.}

We refer to the reference~\cite{masri2023modelling} for a complete model derivation and analysis.

>>>>

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(see also \Cref{fig:intracranial_domains}). The coordinates in these
3D domains is denoted by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
towards the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

In addition, we consider two sets of perivascular networks represented
as graphs: a periarterial network $\Lambda_a$ represented by the
(connected) centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a
perivenous network $\Lambda_v$ associated with the centerlines
$\Lambda^i_{(v)}$ of the veins. Recall that we represent the vascular
domains as the union of cylindrical vessels of radius $R_1^i$
surrounding the centerlines $\Lambda^i$. Moreover, we consider the
surrounding perivascular spaces as the union of annular cylinders
$\Omega^i$ of inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$
and thus of width $R_2^i - R_1^i$. We denote the outer lateral surface
of the periarterial and perivenous spaces by $\Gamma_a$ and
$\Gamma_v$, respectively. We omit the subscript $a$ or $v$ when
referring to any such network, vessel segment or perivascular outer
surface. We assume that $\Lambda$ is parametrized by the coordinate
$s$, and with a minor abuse of notation, simply write $s \in \Lambda$
to represent the point ${\bm \lambda}(s)$ on $\Lambda$ corresponding
to $s$.

@@

In this section, we describe the mathematical
model~\eqref{eq:multi_transport}, the associated definitions,
interface and boundary conditions in further detail. We refer to the
main text for explicit parameter values while giving the general, abstract
form here. Recall that we model a concentration field $c = c(x, t)$
for $x \in \Omega$ and $t > 0$ defined in $\Omega_{\rm PAR}$ and
$\Omega_{\rm CSF}$ separately. In each of these domains, $c$ satisfies
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma = f \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivacular networks $\Lambda_a, \Lambda_v$. To summarize,
\[
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\]
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$ if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$, in which case the interaction with $\Omega_{\rm{CSF}}$ is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

Also in~\eqref{eq:3d_pde}, the notation $\overline{c}$ denotes a lateral average of the concentration over the outer perivascular surfaces, defined for each centerline $\Lambda^i$ and each point $s \in \Lambda^i$ by
\begin{equation}
  \overline{c}(s) = \frac{1}{P(s)} \int_{\partial \Theta_2^i(s)} c 
\end{equation}
where $\partial \Theta_2^i(s)$ is the outer boundary of the cross-section $\Theta(s)$ of the PVS $\Omega^i$ at $s$ and $P(s)$ is the perimeter of $\Theta(s)$. Moreover, for both the periarterial and perivenous networks ($\Lambda_a, \Lambda_v$ with outer PVS boundaries $\Gamma_a, \Gamma_v$), the (Dirac delta) term $\delta_\Gamma$ is a distribution concentrated on the outer lateral surfaces of the PVSs. More precisely, it is defined in terms of its \mer{integrated} action on any sufficiently smooth function $v : \Omega \rightarrow \R$ and for each $\Lambda^i \subseteq \Lambda$ with PVS $\Omega^i$ and outer lateral surface $\Gamma^i = \partial \Omega^i_2$ such that
\begin{equation}
  \int_{\Omega} \xi (\overline{c} - \hat c ) \delta_\Gamma (v) =
  \int_{\Lambda} P \xi (\overline{c} - \hat c ) \overline{v}, 
\end{equation}
under the assumption that $\xi$ is constant on each $\partial
\Theta_2^i$.
%% , since:
%% \begin{multline}
%%   \int_{\Omega} \xi (\overline{c} - \hat c ) \delta_\Gamma (v) =
%%   \int_{\Gamma} \xi (\overline{c} - \hat c ) \delta_\Gamma (v) =
%%   \sum_i \int_{\Gamma^i} \xi (\overline{c} - \hat c ) \delta_\Gamma (v) = 
%%   \sum_i \int_{\Lambda^i} \int_{\partial \Theta_2^i} \xi (\overline{c} - \hat c ) \delta_\Gamma (v) = \\
%%   \sum_i \int_{\Lambda^i} \left (  \xi (\overline{c} - \hat c ) \int_{\partial \Theta_2^i} \delta_\Gamma (v) \right ) =
%%   \sum_i \int_{\Lambda^i} P \xi (\overline{c} - \hat c ) \overline{v} =
%%   \int_{\Lambda} P \xi (\overline{c} - \hat c ) \overline{v} .
%% \end{multline}
We refer to the reference~\cite{masri2023modelling} for a complete model derivation and analysis.
%% \mer{MER: The previous was:
%% \begin{equation}
%%   \xi (\overline{c} - \hat c ) \delta_\Gamma (v) = \int_{\Lambda} P \xi ( \overline{c} - \hat c ) \overline{v} .
%% \end{equation}
%% }

In addtion, we model the concentration $\hat{c} = \hat{c}(s, t)$ in
the periarterial and perivenous networks ($s \in \Lambda_a, s \in
\Lambda_v$) which represents the concentration in the perivascular
space averaged over each perivascular cross-section and satisfies:
\begin{equation}
\partial_t (A  \hat c) - \partial_s(\hat D A \partial_s ( \hat c)) +\partial_s(A \hat u \hat c )  +  \xi P (\hat c - \overline{c}) = 0  \qquad \quad \mathrm{in} \quad  \Lambda_a, \Lambda_v ,
\end{equation}
where $A = A(s)$ is defined for each $s \in \Lambda^i$ and each
centerline $\Lambda^i$ with PVS $\Omega^i$ as the area of the PVS
cross-section $\Theta^i(s)$.
